---
title: "Practical assignment health effects"
author: "X. Pouwels"
date: "26-7-2021"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

The overall aim of this assignment is to get you acquainted with measuring and valuing quality of life using different instruments. The assignment is implemented in a shiny app. 

# Instructions
Install the packages `shiny`, `diagram`, and `tidyverse` before running the app. Run the app using the `runGitHub` function from the `shiny` package as described below. To do so, copy the following code in your own R session. Answer the questions and keep these answers for the discussion.

```{r, include = TRUE, eval = FALSE}
# Set up
rm(list = ls())
install.packages("shiny") # install this package if not already done
install.packages("diagram") # install this package if not already done
install.packages("tidyverse") # install this package if not already done

# Load package
library(shiny)
library(diagram)
library(tidyverse)

# Run shiny app
shiny::runGitHub("Teaching", 
                 "Xa4P", 
                 subdir = "Basics/shiny_app_health_effects/", 
                 ref = "main")
```